Lens Modeling in \sw

### 2013-12-12 text/tex/results.tex
editing sec3
editing sec3

more edits

editing sec3 - added proper def of theta_e_sim, theta_e

editing sec3

The sims were produced using {\tt gravlens}
\citep{2001astro.ph..2341K,2001astro.ph..2340K}.  The were of three
kinds, as follows.

\begin{enumerate}
  \item Imitating lensed quasars: having a singular elliptical
    isothermal lens (SIE) plus constant external shear, and a circular
    Gaussian source.
  \item Emulating lensed galaxies: similar to the above, but with an
    elliptical de Vaucouleurs source.
  \item Resembling cluster lenses: having a source as above, but a
    more complicated lens, with one dominant elliptical SIE and and
    one or more perturbing elliptical SIEs, plus a circular NFW
    \citep{1996ApJ...462..563N,1997ApJ...490..493N} to represent
    the intergalactic mass.
\end{enumerate}

@@

Figure \ref{fig:6941} shows a simple example.  The top two panels show
contours of convergence $\kappa(x,y)$, and of the arrival-time for a
point source at the brightest point of the sim
source.\footnote{Although the arrival-time contours represent an
  abstract quantity that is not directly observable, the contour map
  may actually resemble the appearance of the lensed arcs.  The
  resemblance is due to a serendipitous computer-graphical effect
  \citep{2001AJ....122..585S}.} The spatial scale is in pixels.  The
information in these two panels was, of course, kept secret during the
modelling challenge.  The panels in the middle row and at the lower
right are the results that \spl returns to the modeller, in order for
them to assess the model.  These derive from the mean of an ensemble
of 200 models generated by \spl.  The model mass distribution is
returned as a contour map superimposed on an intensity map. A fairly
smooth mass distribution, as here, is a good sign.  An irregular or
checkboard pattern in the mass map usually indicates a bad model.  The
model arrival-time contours.  The special contour passing through a
saddle point is shown in black: this is the model version of the
spaghetti sketch provided by the user.  Also provided is a synthetic
image of the lensed source.  For this, \spl assumes a simple conical
source profile.  The user can change the contrast level on the image,
which (though it is not saved) amounts to to adjusting the width of
the cone.  Finally, to the lower left of the figure, we have a
comparison of the input and recovered mass profiles.  The panel shows
a average $\kappa$ within a given radius, as a function of radius.
The red curve is the true value, and where it crosses unity (dotted
horizonal line) is the notional Einstein radius.  The two blue curves
are the minimal and maximal mean enclosed $\kappa$ from the internal
ensemble in \spl.  The region between the blue curves is shaded
between the radii of the innermost and the outermost images --- this
is the confidence region from the modelling.  As we see, the shaded
blue is slightly above the red curve.  In summary: the identification
of minimum and saddle point is correct, but the estimated Einstein
radius is a little too high.

@@

T1 tested the volunteers ability to reconstruct the arrival time surface given a survey image containing a sim.
This task consists of two parts.
First, the correct identification and location of lensed images (T1a).
Second the correct ordering for the identified lensed images in respect of the arrival time (T1b).

While we expected T1a to be trivial, given the nature of the survey images and the success of \sw, we expect T1b to be more difficult.
T1b tests the volunteers understandings of the theory of arrival time surfaces and the odd number theorem.
While we can provide the volunteers with some general rules of thumb, T1b involves imagination and guessing and therefore training could improve their skills in a later stage.

T1 was also designed to give some feedback on the difficulties volunteers encounter, to further improve the tutorial materials.

The evaluation of the volunteers performance for T1 was done manually, comparing their input from \spl and the resulting reconstructed arrival time surface contour line plot (arrival plot) to the arrival plot generated using simulation parameters.
% \Figref{output_compare} shows the setup used to evaluate the models.

To evaluate T1a and T1b, each model was evaluated with two binary tests.
The first is passed, if all the images have been identified and are approx. within $\pm0.05\cdot\text{imgage width}$.
The second test is passed, if those identified points have the right parity, ordering with respect to arrival time.

Additionally, ten types of errors (E01 -- E10, listed in \tabref{stats}) that could occur were analyzed, where each model could contain multiple of those.

@@

In \tabref{stats} a summary of this evaluation is presented.
We conclude that the volunteers are performing very well identifying and positioning images (T1a), with a performance of 92\% (R1, p=0.92).
Most of the problems where due to unclear arc-like structures (E01, p=0.18; E04, p=0.03; E08, p=0.04).
Critical errors like the failure to identify all five images in a five images system (E03, p=0.04) or to include too many images (E10, p=0.01) did almost never happen.
From this we conclude that the introduction materials was adequate and the volunteers understand the basics of gravitational lensing.

\todo{add more detail?} Add more details like total number of images detected / tot am images (rightPlace fraction)??

The assignment of the parity of the images (T1b) was a more difficult task.
In 59\% (R2, p=0.59, N=70) of the cases the volunteers succeeded to identify the right configuration.
Most of the failures are due to E06, with N=38, p=32\%, followed by E05 (PI) with N=7, p=6\%.
For an example of E06, see \figref{7022}.
It basically describes a situation, where the minima and saddle points of a five image configuration were exchanged (rotated by $\pm90\dgr$).
E05 describes the situation, where the ordering of the saddle points was wrong (rotation by $180\dgr$).
While these errors occurred, we suspect they can be avoided with better training material and some examples for the obvious cases.
For more challenging cases, like very symmetrical distribution of the lensed images (for example model 7022, \figref{7022}), those errors should still produce plausible results, as will be explained in the next section.

@@

The second test T2 was to compare the desired results of a modeling process, the mass distribution of the lens $\kappa(x, y)$.
To get a means of comparing the simulated data (sims) to the generated modeled data (models), the total convergence, also called enclosed mass $\kappa_{\text{encl}}(r)$ for both was calculated.
The Einstein radius $\Theta_E$ is defined by $\kappa_{\text{encl}}(\Theta_E)=1$ and gives a number that allows the comparison between the sims and the models.
We also let an expert model three selected systems to compare the results from volunteers to those of a professional.

\todo{!} Should I write down prior expectation for T2 too?

To get a means of comparing the simulated data (sims) to the generated modeled data (models), the total convergence, also called enclosed mass $\kappa_{\text{encl}}(r)$ for both was calculated.
The Einstein radius $\Theta_E$ is defined by $\kappa_{\text{encl}}(\Theta_E)=1$ and gives a number that allows the comparison between the sims and the models.
We also let an expert model three selected systems to compare the results from volunteers to those of a professional.

\todo{!} Should I write down prior expectation for T2 too?

To compare the enclosed mass profile and the Einstein radius of the simulation and the models, \kenc was calculated using the mass map \kap[x,y] directly generated in the modeling process.
From the ensemble of models generated by one modeling process, the mean is taken as the resulting $\kappa(r)$ to calculate $\theta_E$.
To estimate the errors, also the extremal models are used to estimate a lower and upper limit for $\theta_E$.
These results can be seen in \figref{ER_all_models}.
This figure shows that this technique of estimating the error minimizes the error significantly and should be improved for further analysis.

In \figref{ER_per_sim} can be seen, that the calculated \ERf of the models tend to be too high.
The overshoot varies from around 0.2 to 0.4 for good models.
One of the reasons for this is, that it's hard to get the center of the lens on spot.
An offset leads to a a flatter mass profile for the model compared to the simulation.

>>>>

The sims were produced using {\tt gravlens}
\citep{2001astro.ph..2341K,2001astro.ph..2340K}.  There were of three
kinds, as follows.

\begin{enumerate}
  \item Imitating lensed quasars: having a singular elliptical
    isothermal lens (SIE) plus constant external shear, and a circular
    Gaussian source.
  \item Emulating lensed galaxies: similar to the above, but with an
    elliptical de Vaucouleurs source.
  \item Resembling cluster lenses: having a source as above, but a
    more complicated lens, with one dominant elliptical SIE and
    one or more perturbing elliptical SIEs, plus a circular NFW
    \citep{1996ApJ...462..563N,1997ApJ...490..493N} to represent
    the intergalactic mass.
\end{enumerate}

@@

Figure \ref{fig:6941} shows a simple example.  The top two panels show
contours of convergence $\kappa(x,y)$, and of the arrival-time for a
point source at the brightest point of the sim
source.\footnote{Although the arrival-time contours represent an
  abstract quantity that is not directly observable, the contour map
  may actually resemble the appearance of the lensed arcs.  The
  resemblance is due to a serendipitous computer-graphical effect
  \citep{2001AJ....122..585S}.} The spatial scale is in pixels.  The
information in these two panels was, of course, kept secret during the
modelling challenge.  The panels in the middle row and at the lower
right are the results that \spl returns to the modeller, in order for
them to assess the model.  These derive from the mean of an ensemble
of 200 models generated by \spl.  The model mass distribution (left mid) is
returned as a contour map superimposed on an intensity map. A fairly
smooth mass distribution, as here, is a good sign.  An irregular or
checkboard pattern in the mass map usually indicates a bad model.  The
models arrival-time contours are shown in the right middle panel.
  The special contour passing through a
saddle point is shown in black: this is the model version of the
spaghetti sketch provided by the user.  Also provided is a synthetic
image of the lensed source in the bottom right panel.
  For this, \spl assumes a simple conical
source profile.  The user can change the contrast level on the image,
which (though it is not saved) amounts to adjusting the width of
the cone.  Finally, to the lower left of the figure, we have a
comparison of the input and recovered mass profiles.  The panel shows
a average $\kappa$ within a given radius, as a function of radius.
The red curve is the true value, and where it crosses unity (dotted
horizonal line) is the notional Einstein radius $\Theta_{\text{E, sim}}$.
  The two blue curves
are the minimal and maximal mean enclosed $\kappa$ from the internal
ensemble in \spl.  The region between the blue curves is shaded
between the radii of the innermost and the outermost images --- this
is the confidence region from the modelling.  As we see, the shaded
blue is slightly above the red curve. 
The Einstein Radius $\Theta_\text{E}$ of the model is estimated crossing the
mean enclosed $\kappa$ (not plotted) with unity.
  In summary: the identification
of minimum and saddle point is correct, but the estimated Einstein
radius is a little too high.

@@

A first manual evaluation tested the volunteers ability to reconstruct the arrival time surface given a survey image containing a sim.
This task consists of two parts.
First to correctly identify and locate the lensed images.
Second the correct ordering for the identified lensed images in respect of the arrival time.

While we expected the identification of lensed images to be trivial, given the nature of the survey images and the success of \sw, we expect the correct ordering to be more difficult.
This tests the volunteers understandings of the theory of arrival time surfaces and the odd number theorem.
While we can provide the volunteers with some general rules of thumb, ordering involves imagination and guessing and therefore training could improve results in a later stage.

This first test was designed to give some feedback on the difficulties volunteers encounter, to further improve the tutorial materials.

The evaluation of the volunteers performance was done manually, comparing their input from \spl and the resulting reconstructed arrival time surface contour line plot (arrival plot) to the arrival plot generated using simulation parameters.
% \Figref{output_compare} shows the setup used to evaluate the models.

The images of the system are considered to be identified correcly, if all the images have been identified and are approx. within $\pm0.05\cdot\text{imgage width}$.
The parity is considered correct, if those identified points have the right ordering with respect to arrival time.

Additionally, ten types of errors (labeled E01 -- E10, listed in \tabref{stats}) that occurred were identified.
Each generated model could contain more than one error.

@@

\tabref{stats} presents a summary of this evaluation.
We conclude that the volunteers are performing very well identifying and positioning images, with a performance of 92\% (R1, p=0.92).
Most of the problems where due to unclear arc-like structures (E01, p=0.18; E04, p=0.03; E08, p=0.04).
Critical errors like the failure to identify all five images in a five images system (E03, p=0.04) or to include too many images (E10, p=0.01) did almost never happen.
From this we conclude that the introduction materials was adequate and the volunteers understand the basics of gravitational lensing.

The assignment of the parity of the images was a more difficult task.
In 59\% (R2, p=0.59, N=70) of the cases the volunteers succeeded to identify the right configuration.
Most of the failures are due to E06 (N=38, p=32\%), followed by E05 (N=7, p=6\%).
E06 describes a situation, where the minima and saddle points of a five image configuration were exchanged (rotated by $\pm90\dgr$), see \figref{7022} for an example.
E05 describes a situation, where the ordering of the saddle points was wrong (rotation by $180\dgr$).
While these errors occurred, we suspect they can be avoided with better training material and some examples for the obvious cases.
For more challenging cases, like very symmetrical distribution of the lensed images (for example model 7022, \figref{7022}), those errors should still produce plausible results, as will be explained in the next section.

@@

The second test was to compare the mass distribution of the lens $\kappa(x, y)$ of the models and the sims.
To get a means of comparing the sims to the models, the total convergence, also called enclosed mass $\kappa_{\text{encl}}(r)$ was calculated for both.
The Einstein radius $\Theta_E$ is defined by $\kappa_{\text{encl}}(\Theta_E)=1$ and gives a number that allows the comparison between a sim and a model.
We also let an expert (PS) model three selected systems to compare the results from volunteers to those of a professional.

To compare the enclosed mass profile and the Einstein radius of the simulation and the models, \kenc was calculated using the mass map \kap[x,y] generated in the modeling process.
From the ensemble of models generated by one modeling process, the mean is taken as the resulting $\kappa(r)$ to calculate $\theta_E$.
To estimate the errors, also the extremal models are used to estimate a lower and upper limit for $\theta_E$.
These results can be seen in \figref{ER_all_models}.
This figure shows that this technique of estimating the error minimizes the error significantly and should be improved for further analysis.

In \figref{ER_per_sim} can be seen, that the calculated \ERf of the models tend to be too high.
The overshoot varies from around 0.2 to 0.4 for good models.
One of the reasons for this is, that it's hard to get the center of the lens on spot.
An offset leads to a flatter mass profile for the model compared to the simulation.

### 2013-12-12 text/tex/theory.tex
editing sec3
editing sec3

more edits

editing sec3 - added proper def of theta_e_sim, theta_e

editing sec3

The geometrical time delay \eqref{eq:Ageom} is not the only one.  The
warping of space by a gravitational field introduces a further time
delay $A_{\rm grav}$.  The arrival time of the light ray, compared to
what it would have been with no lensing, is
\begin{equation} \label{eq:Aarriv}
\hbox{arrival time} = A_{\rm geom} + A_{\rm grav} \,.
\end{equation}
We can also express $A_{\rm grav}$ as an area, with the same implicit
constant factor that makes it a time.  The expression, however, is
more complicated.  We first introduce the notation $\nabla^2 f(x,y)$
to denote
\begin{equation}
 \frac{ f(x+\Delta x, y) + f(x-\Delta x, y) +
        f(x, y+\Delta y) + f(x, y-\Delta y) - 4 f(x,y) }
      {\Delta x \; \Delta y}
\end{equation}
in the limit of $\Delta x,\Delta y$ small.  Then we write the
projected mass density of the lens, which would be in $\rm
kg\,m^{-2}$, in special units (given in the Appendix) such that the
surface density becomes a dimensionless field $\kappa(x,y)$.  Then
\begin{equation} \label{eq:Poisson}
\nabla^2 A_{\rm grav}(x,y) = -2\kappa(x,y) \,.
\end{equation}
Note that $A_{\rm grav}(x,y)$ is not given explicitly in terms of
$\kappa(x,y)$, but as a differential equation that must be solved.
Equations of the type \eqref{eq:Poisson} are, however, well known (as
Poisson's equation in two dimensions), and there are techniques for
solving them.  Incidentally, $\kappa$ has a second meaning, as well as
being a dimensionless form of the the projected density: it also
measures how a bundle of light rays is brought together by
gravitational lensing, and is called the {\em convergence}.

>>>>

The geometrical time delay \eqref{eq:Ageom} is not the only one.  The
warping of space by a gravitational field introduces a further time
delay $A_{\rm grav}$.  The arrival time of the light ray, compared to
what it would have been with no lensing, is
\begin{equation} \label{eq:Aarriv}
\hbox{arrival time} = A_{\rm geom} + A_{\rm grav} \,.
\end{equation}
We can also express $A_{\rm grav}$ as an area, with the same implicit
constant factor that makes it a time.  The expression, however, is
more complicated.  We first introduce the notation $\nabla^2 f(x,y)$
to denote
\begin{equation}
 \frac{ f(x+\Delta x, y) + f(x-\Delta x, y) +
        f(x, y+\Delta y) + f(x, y-\Delta y) - 4 f(x,y) }
      {\Delta x \; \Delta y}
\end{equation}
in the limit of $\Delta x,\Delta y$ small.  Then we write the
projected mass density of the lens, which would be in $\rm
kg\,m^{-2}$, in special units (given in the Appendix) such that the
surface density becomes a dimensionless field $\kappa(x,y)$.  Then
\begin{equation} \label{eq:Poisson}
\nabla^2 A_{\rm grav}(x,y) = -2\kappa(x,y) \,.
\end{equation}
Note that $A_{\rm grav}(x,y)$ is not given explicitly in terms of
$\kappa(x,y)$, but as a differential equation that must be solved.
Equations of the type \eqref{eq:Poisson} are, however, well known (as
Poisson's equation in two dimensions), and there are techniques for
solving them.  Incidentally, $\kappa$ has a second meaning, as well as
being a dimensionless form of the projected density: it also
measures how a bundle of light rays is brought together by
gravitational lensing, and is called the {\em convergence}.
